The Minimal Retroreflective Microfacet Model

### 2025-07-10 article.tex
- add illustration - unify math - add Fresnel to plausibility - updates on text

% Once an article is accepted accepted, switch to the following line and comment the preceding one. The editor will supply the argument values.
\accepted{yyyy-mm-dd}{yyyy-mm-dd}{yyyy-mm-dd}{Editor Name}{volume}{issue}{1}{6}{yyyy}
\seturl{http://jcgt.org/published/vol/issue/num/}

\newcommand{\vecv}{\mathbf{v}}
\newcommand{\vecvv}{\mathbf{v'}}
\newcommand{\vecl}{\mathbf{l}}
\newcommand{\vecll}{\mathbf{l'}}
\newcommand{\vecn}{\mathbf{n}}
\newcommand{\vecm}{\mathbf{m}}
\newcommand{\vech}{\mathbf{h}}
\newcommand{\vecb}{\mathbf{b}}
\newcommand{\vdot}[2]{\left\langle #1, #2 \right\rangle}

@@

In this document, we detail the correct usage and derivation of this retro-reflective BRDF model within the microfacet framework~\cite{WalterMicrofacet,HeitzMicrofacet}. We first recall some basic principles from microfacet theory (Section~\ref{sec:microfacets}) and show how to account for retro-reflective microfacets (Section~\ref{sec:backvector}) within this framework. We then detail how to implement this in practice (Section~\ref{sec:implementation}). Finally, we discuss how retro-reflection can occur at the microfacet level (Section~\ref{sec:plausibility}).

@@

\begin{figure}
  \begin{center}
    \begin{tabular} {|p{0.2\textwidth}|p{0.7\textwidth}|}
%    \begin{tabular} {|l|l|}
      \hline
      $\vecv$ & normalized vector to viewer \\
      \hline
      $\vecl$ & normalized vector to light \\
      \hline
      $\vecn$ & surface normal \\
      \hline
      $\mathbf{t_1}, \mathbf{t_2}$ & tangent and bitangent (forming an orthonormal coordinate frame with the normal) \\
      \hline
      $\vech$ & half vector, $\vech = \vech(\vecv, \vecl) = \frac{\vecv + \vecl}{\lVert \vecv + \vecl \rVert}$ \\
      \hline
      $\vdot{\mathbf{x}}{\mathbf{y}}$ & dot product of $\mathbf{x}$ and $\mathbf{y}$ \\
      \hline
      $\text{reflect}(\mathbf{x}, \mathbf{m})$ & reflection of vector $\mathbf{x}$ on surface with normal $\mathbf{m}$, i.e. $\text{reflect}(\mathbf{x}, \mathbf{m}) = -\mathbf{x} + 2 \vdot{\mathbf{x}}{\mathbf{m}} \mathbf{m}$ \\
      \hline
      $\Omega$ & hemisphere of directions around the normal \\
      \hline
    \end{tabular}
  \end{center}
  \caption{Notations used.}
  \label{fig:notation}
\end{figure}

@@

Instead of half-vector $\vech$, we can use the back-vector $\vecb$~\cite{BelcourBackvector} defined as
\begin{equation}
  \vecb(\vecv, \vecl) = \vech(\vecvv, \vecl) = \frac{\vecvv + \vecl}{\lVert \vecvv + \vecl \rVert}, \text{ with } \vecvv = \text{reflect}(\vecv, \vecn),
\end{equation}
to select the reflecting microfacet. %, we are to re-derive the microfacet model.
As illustrated in figure \ref{fig:backvector}, the reflection $\vecvv$ of the view direction $\vecv$ on the surface can also be interpreted as an reflection of the view direction on a plane aligned with $\vecv$ and perpendicular to the surface.
Intuitively, this is equivalent of observing the microfacet reflection through a virtual mirror, thus turning glossy forward-reflection into glossy retro-reflection.
While there is no immediate physical foundation to this construction, this interpretation can serve as motivation why components from the regular microfacet model still work when changing the domain to the back-vector.
\begin{figure}
  \begin{center}
    \includegraphics[width=0.2\textwidth]{figures/sphere_front_r0.3.png} % TODO: add sketch
  \end{center}
  \caption{Geometry of the back vector, illustrated in two dimensions.}
  \label{fig:backvector}
\end{figure}

\paragraph*{Jacobian of Back-Vector.}
First we observe that  Jacobian of the change of variable from the back-vector to the light vector $\vecl$~(Equation~(11) and~(14) of \citet{WalterMicrofacet}) has the same form as the half-vector one:
\begin{equation}
\left\lVert \dfrac{\partial \omega_{\vecb}}{\partial \omega_{\vecl}} \right\rVert = \left\lVert \dfrac{\partial \omega_{\vech(\vecvv, \vecl)}}{\partial \omega_{\vecl}} \right\rVert = \dfrac{1}{4 \left| \vecl \cdot \vech(\vecvv, \vecl) \right|} = \dfrac{1}{4 \left| \vecl \cdot \vecb \right|}
\end{equation}

\paragraph*{Microfacet reflection term.}
There is experimental evidence~\cite{BelcourBackvector} that using a Fresnel term based on the back-vector can provide a good match to actual retro-reflective material measurements.
As such, we update the formulation of $\rho(\vecv, \vecm)$~(Equation (11) of \citet{WalterMicrofacet}) to use the reflected view vector instead. It follows that the reflection microsurface BRDF~(Equation (15) of \citet{WalterMicrofacet}) becomes
\begin{equation}
  f_r^m(\vecv, \vecl, \vecm) = F(\vecvv, \vecm) \dfrac{\delta(\vecb, \vecm) }{4 \left| \vecvv \cdot \vecb \right|^2}.
\end{equation}
We used the fact that $\left| \vecl \cdot \vecb \right| = \left| \vecvv \cdot \vecb \right|$.

@@

\paragraph*{Final form.}
The final retro-reflective microfacet model can be written as
\begin{equation}
  f_\text{retro}(\vecv, \vecl) = f(\vecvv, \vecl) = \frac{D(\vecb) \, G_{2}(\vecvv,\vecl) \, F(\vecb, \vecvv)}{4 \vdot{\vecvv}{\vecn} \vdot{\vecl}{\vecn}}
  \label{eq:retro_microfacet_model}.
\end{equation}
In the following section we show in detail that this model is plausible, i.e. that symmetry and energy conservation requirements are fulfilled.

@@

As anisotropic Phong, Beckmann, and GGX distributions are reflection-symmetric around the normal, i.e. they take the form $D(\vech) = D\left(\vdot{\vecn}{\vech}, \lvert \vdot{\mathbf{t_1}}{\vech} \rvert, \lvert \vdot{\mathbf{t_2}}{\vech} \rvert\right)$, we can thus infer that using them with the back-vector obeys symmetry.

@@

@@

Using this along with equations \eqref{eq:nh} and \eqref{eq:nk} we get

@@

  \frac{D(\vech(\vecvv, \vecl)) G_{2}(G_1(\vecvv, \vecl), G_1(\vecl, \vecvv))}{4 \vdot{\vecvv}{\vecn} \vdot{\vecl}{\vecn} }\\

@@

  \frac{D(\vech(\vecll, \vecv)) G_{2}(G_1(\vecv, \vecll), G_1(\vecll, \vecv))}{4 \vdot{\vecv}{\vecn} \vdot{\vecll}{\vecn} } =

@@

\begin{itemize}
\item
  Evaluation merely needs to replace $\vecv$ with $\vecvv$ upfront.
\item
  Similarly, importance sampling of $\vecl$ given $\vecv$ can be realized by replacing $\vecv$ with $\vecvv$ upfront and then importance sampling the regular microfacet BRDF.
  This may include low variance sampling using the domain of visible microfacets \cite{HeitzIS}.
\item
  As the albedos of standard BRDF and retro-reflective BRDF are essentially identical, compensating for energy loss in the sense of \cite{KelemenBRDF} can be realized using the same data tables.
\item Pseudo-code is provided in Listing~\ref{lst:implementation}.
\end{itemize}

>>>>

% Once an article is accepted accepted, switch to the following line and comment the preceding one. The editor will supply the argument values.
\accepted{yyyy-mm-dd}{yyyy-mm-dd}{yyyy-mm-dd}{Editor Name}{volume}{issue}{1}{6}{yyyy}
\seturl{http://jcgt.org/published/vol/issue/num/}
\newcommand{\vecv}{\mathbf{v}}
\newcommand{\vecvv}{\mathbf{v'}}
\newcommand{\vecl}{\mathbf{l}}
\newcommand{\vecll}{\mathbf{l'}}
\newcommand{\vecn}{\mathbf{n}}
\newcommand{\vecm}{\mathbf{m}}
\newcommand{\vech}{\mathbf{h}}
\newcommand{\vecb}{\mathbf{b}}
\newcommand{\vecx}{\mathbf{x}}
\newcommand{\vecy}{\mathbf{y}}
\newcommand{\vdot}[2]{\left\langle #1, #2 \right\rangle}

@@

In this document, we detail the correct usage and derivation of this retro-reflective BRDF model within the microfacet framework~\cite{WalterMicrofacet,HeitzMicrofacet}.
We first recall some basic principles from microfacet theory (Section~\ref{sec:microfacets}), then show how this model can be steered from forward- to retro-reflection (Section~\ref{sec:backvector}), and further provide proof that the resulting BRDF is fully plausible (Section~\ref{sec:plausibility}).
Finally, we sketch how the model can be implemented in practice (Section~\ref{sec:implementation}).

@@

\begin{figure}
  \begin{center}
    \begin{tabular} {|p{0.2\textwidth}|p{0.7\textwidth}|}
%    \begin{tabular} {|l|l|}
      \hline
      $\vecv$ & normalized vector to viewer \\
      \hline
      $\vecl$ & normalized vector to light \\
      \hline
      $\vecn$ & surface normal \\
      \hline
      $\mathbf{t_1}, \mathbf{t_2}$ & tangent and bitangent (forming an orthonormal coordinate frame with the normal) \\
      \hline
      $\vech$ & half vector, $\vech = \vech(\vecv, \vecl) = \frac{\vecv + \vecl}{\lVert \vecv + \vecl \rVert}$ \\
      \hline
      $\vdot{\vecx}{\vecy}$ & dot product of $\vecx$ and $\vecy$ \\
      \hline
      $\text{reflect}(\mathbf{x}, \mathbf{m})$ & reflection of vector $\mathbf{x}$ on surface with normal $\mathbf{m}$, i.e. $\text{reflect}(\mathbf{x}, \mathbf{m}) = -\mathbf{x} + 2 \vdot{\mathbf{x}}{\mathbf{m}} \mathbf{m}$ \\
      \hline
      $\Omega$ & hemisphere of directions around the normal \\
      \hline
    \end{tabular}
  \end{center}
  \caption{Notations used.}
  \label{fig:notation}
\end{figure}

@@

Instead of half-vector $\vech$, we can use the back-vector $\vecb$~\cite{BelcourBackvector} defined as
\begin{equation}
  \vecb(\vecv, \vecl) = \vech(\vecvv, \vecl) = \frac{\vecvv + \vecl}{\lVert \vecvv + \vecl \rVert}, \text{ with } \vecvv = \text{reflect}(\vecv, \vecn),
\end{equation}
to select the reflecting microfacet. %, we are to re-derive the microfacet model.

\begin{figure}[h]
  \begin{center}
    \includegraphics[width=0.4\textwidth]{figures/backvector.pdf}
  \end{center}
  \caption{Geometry of the back vector, illustrated in two dimensions.}
  \label{fig:backvector}
\end{figure}

As illustrated in figure \ref{fig:backvector}, the reflection $\vecvv$ of the view direction $\vecv$ on the surface can also be interpreted as an reflection of the view direction on a plane aligned with $\vecv$ and perpendicular to the surface.
Intuitively, this is equivalent of observing the microfacet reflection through a virtual mirror, thus turning glossy forward-reflection into glossy retro-reflection.
While there is no immediate physical foundation to this construction, this interpretation can serve as motivation why components from the regular microfacet model still work when changing the domain to the back-vector.

\paragraph*{Jacobian of Back-Vector.}
First we observe that  Jacobian of the change of variable from the back-vector to the light vector $\vecl$~(Equation~(11) and~(14) of \citet{WalterMicrofacet}) has the same form as the half-vector one:
\begin{equation}
\left\lVert \dfrac{\partial \omega_{\vecb}}{\partial \omega_{\vecl}} \right\rVert = \left\lVert \dfrac{\partial \omega_{\vech(\vecvv, \vecl)}}{\partial \omega_{\vecl}} \right\rVert = \dfrac{1}{4 \left| \vdot{\vecl}{\vech(\vecvv, \vecl)} \right|} = \dfrac{1}{4 \left| \vdot{\vecl}{\vecb} \right|}
\end{equation}

\paragraph*{Microfacet reflection term.}
There is experimental evidence~\cite{BelcourBackvector} that using a Fresnel term based on the back-vector can provide a good match to actual retro-reflective material measurements.
As such, we update the formulation of $\rho(\vecv, \vecm)$~(Equation (11) of \citet{WalterMicrofacet}) to use the reflected view vector instead. It follows that the reflection microsurface BRDF~(Equation (15) of \citet{WalterMicrofacet}) becomes
\begin{equation}
  f_r^m(\vecv, \vecl, \vecm) = F(\vecvv, \vecm) \dfrac{\delta(\vecb, \vecm) }{4 \left| \vdot{\vecvv}{\vecb} \right|^2}.
\end{equation}
We used the fact that $\left| \vdot{\vecl}{\vecb} \right| = \left| \vdot{\vecvv}{\vecb} \right|$.

@@

\paragraph*{Final form.}
The final retro-reflective microfacet model can be written as
\begin{equation}
  f_\text{retro}(\vecv, \vecl) = f(\vecvv, \vecl) = \frac{D(\vecb) \, G_{2}(\vecvv,\vecl) \, F(\vecvv, \vecb)}{4 \vdot{\vecvv}{\vecn} \vdot{\vecl}{\vecn}}
  \label{eq:retro_microfacet_model}.
\end{equation}
In the following section we show in detail that this model is plausible, i.e. that symmetry and energy conservation requirements are fulfilled.
Note that for this we impose one technical assumption on $D$:
we require that the microfacet normal distribution is reflection-symmetric around the normal (which is the case for all isotropic distributions and for anisotropic Phong, Beckmann, and GGX).

@@

As $D$ is reflection-symmetric around the normal, i.e. it takes the form $D(\vech) = D\left(\vdot{\vecn}{\vech}, \lvert \vdot{\mathbf{t_1}}{\vech} \rvert, \lvert \vdot{\mathbf{t_2}}{\vech} \rvert\right)$, we can thus infer that using it with the back-vector obeys symmetry.

@@

\end{equation}\label{eq:F}
As the Fresnel term takes the form $F(\vecx, \vecy) = F\left(\vdot{\vecx}{\vecy}\right)$, we can conclude
    F\left((\vecvv, \vech(\vecvv, \vecl)\right) = F\left(\vdot{\vecvv}{\vech(\vecvv, \vecl)}\right) =
    F\left(\vdot{\vecll}{\vech(\vecll,\vecv)}\right) = F\left(\vecll, \vech(\vecll,\vecv)\right).

@@

Further, using equation \eqref{eq:kh} along with equations \eqref{eq:nh} and \eqref{eq:nk} we get

@@

  \frac{D(\vech(\vecvv, \vecl)) G_{2}(G_1(\vecvv, \vecl), G_1(\vecl, \vecvv)) F(\vecvv, \vecl)}{4 \vdot{\vecvv}{\vecn} \vdot{\vecl}{\vecn} }\\

@@

  \frac{D(\vech(\vecll, \vecv)) G_{2}(G_1(\vecv, \vecll), G_1(\vecll, \vecv)) F(\vecll, \vecv)}{4 \vdot{\vecv}{\vecn} \vdot{\vecll}{\vecn} }\\

@@

\begin{itemize}
\item
  Evaluation merely needs to replace $\vecv$ with $\vecvv$ upfront.
\item
  Similarly, importance sampling of $\vecl$ given $\vecv$ can be realized by replacing $\vecv$ with $\vecvv$ upfront and then importance sampling the regular microfacet BRDF.
  This may include low variance sampling using the domain of visible microfacets \cite{HeitzIS}.
\item
  As the albedos of standard BRDF and retro-reflective BRDF are essentially identical, compensating for energy loss in the sense of \cite{KelemenBRDF} can be realized using the same data tables.
\item Pseudo-code is provided in Listing~\ref{lst:implementation}.
  \todo{mraab: I think source code may be overkill, given the simplicity of the model.}
\end{itemize}
